---
title: "College Basketball Analysis"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
output: html_document
---
T
```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
library(magrittr)
```
This code below loads the untouched dataframe then creates a new dataframe used to clean up the data. For the 
most part, the data is already tidy, with a team's performance in one season being an entity. Here, we add a column
which we will fill out later assigning an integer value to a team's postseason performance ("pstsn_rank"). We also
add an attribute called win_ratio, which computes the wins to games played ratio.
```{r initialize dataframe}
bball_tab <- read_csv("~/CMSC320-Final-Project/cbb.csv", col_types = cols())
clean_bball <- bball_tab %>% 
  filter(!is.na(`POSTSEASON`)) %>%
  mutate(pstsn_rank=0) %>%
  mutate(entityid=row_number()) %>%
  mutate(win_ratio=W/G)
```

This section takes the text feature of the original table and assigns a number to the 
```{r clean_data}
for (i in 1:340) {
  j<-clean_bball[i,"POSTSEASON"]
  if(j=="Champions"){
    clean_bball[i,"pstsn_rank"]<-7
  } else if(clean_bball[i,"POSTSEASON"]=="2ND"){
    clean_bball[i,"pstsn_rank"]<-6
  } else if(clean_bball[i,"POSTSEASON"]=="F4"){
    clean_bball[i,"pstsn_rank"]<-5
  } else if(clean_bball[i,"POSTSEASON"]=="E8"){
    clean_bball[i,"pstsn_rank"]<-4
  } else if(clean_bball[i,"POSTSEASON"]=="S16"){
    clean_bball[i,"pstsn_rank"]<-3
  } else if(clean_bball[i,"POSTSEASON"]=="R32"){
    clean_bball[i,"pstsn_rank"]<-2
  } else if(clean_bball[i,"POSTSEASON"]=="R64"){
    clean_bball[i,"pstsn_rank"]<-1
  } else {
    clean_bball[i,"pstsn_rank"]<-0
  }
}

ncaa_tab <- clean_bball[,c(26,27,25,23,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,24)]
ncaa_tab <- subset(ncaa_tab, select = -c(TEAM,CONF,G,W,POSTSEASON,YEAR))
```


